---
title: "Bacterial Genomics - Summary Report"
date: today
author: 
  - name: "Marcela Torres <br> Dahlia Walters"
    affiliation: "Molecular Epidemiology Program, WA DOH"
format: 
  html:
    self-contained: true
params:
  taxa: "Placeholder"
  phylogeny: "all" #defaults to showing both Snippy and Gubbins results but the rendering script will override this if Gubbins or Snippy are selected
  PatientAddressCounty: "all"  #defaults to showing all samples regardless of the patient address county
  epiprogram: "na"  #defaults to not showing healthcare facility of origin name in the first metadata table if HAI is not specified
---

```{r, results='hide', echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(tidyverse)
library(lubridate)
library(tools)
library(htmltools)
library(fs)
library(rjson)
```

```{r, results='hide', echo=FALSE, message=FALSE, warning=FALSE}
source("scripts/load_files.R")
source("scripts/summarytsv_outputs.R")
source("scripts/SNP_min_max.R")
source("scripts/metadata_summarytsv.R") 
source("scripts/addwaid_microreact.R")
source("scripts/partition_info.R")
```

```{r taxa, echo = FALSE}
# Load the taxa dataset
results_by_taxa <- read.csv("outputs_scripts/results_by_taxa.csv", stringsAsFactors = FALSE)

# Filter for the specific taxa
taxa_data <- results_by_taxa %>%
  filter(TAXA == params$taxa)

# Check if data is available for this taxa
if (nrow(taxa_data) == 0) {
  cat("No data available for this taxa.")
  knitr::knit_exit()  # Stops execution for empty reports
}
```


```{r phylogeny, echo = FALSE}
valid_phylogenies <- c("Snippy", "Gubbins", "all")

if (!(params$phylogeny %in% valid_phylogenies)) {
  stop("Invalid phylogeny parameter. Choose 'Snippy', 'Gubbins', or 'all'.")
}

phylogeny <- params$phylogeny
```


```{r PatientAddressCounty, echo = FALSE}
PatientAddressCounty <- params$PatientAddressCounty

PatientAddressCounty <- toTitleCase(tolower(PatientAddressCounty))
```

```{r epidprogram, echo = FALSE}
epiprogram <- params$epiprogram
```


```{r global helper function, echo = FALSE}
# Load summary_table which has patient address county info
outputs_script_dir <- "outputs_scripts"
summary_table <- file.path(outputs_script_dir, "summary_table.RData")
load(summary_table)

# Create county lookup
county_lookup <- summary_table %>%
  select(ID, PatientAddressCounty)

county_param <- toTitleCase(tolower(params$PatientAddressCounty))

# Global helper function
add_and_filter_county <- function(df) {
  
  # Detect ID column
  id_col <- if ("ID" %in% colnames(df)) {
    "ID"
  } else if ("WA ID" %in% colnames(df)) {
    "WA ID"
  } else {
    return(df)
  }
  
  # Only join if PatientAddressCounty is missing
  if (!"PatientAddressCounty" %in% colnames(df)) {
    df <- df %>%
      left_join(county_lookup, by = setNames("ID", id_col))
  }
  
  county_param <- tools::toTitleCase(tolower(params$PatientAddressCounty))
  
  if (county_param != "all") {
    df <- df %>%
      mutate(PatientAddressCounty = tools::toTitleCase(tolower(PatientAddressCounty))) %>%
      filter(stringr::str_detect(PatientAddressCounty, county_param))
  }
  
  df
}
```


```{r summary_table_display, echo = FALSE}
summary_table <- file.path("outputs_scripts", "summary_table.RData")
load(summary_table)


#Using global helper function filter based on params PatientAddressCounty 
summary_table_filtered <- summary_table %>%
  add_and_filter_county()

#Rename Columns
summary_table_display <- summary_table_filtered  %>%
  rename(
    "WA ID" = ID,
    "ALT ID" = ID_ALT,
    "WDRS ID" = CASE_ID,
    "Collection Date" = SpecimenDateCollected,
    "Patient County" = PatientAddressCounty,
    "Submitter County" = SubmitterCounty,
    "Submitter Facility" = SubmitterName,
    "Origin Facility" = OriginHealthFacility,
    "Specimen Souce"=SpecimenSource
  )

```

## Overview New Samples

In this BigBacter run, there are `r nrow(summary_table_filtered)` new sequencing results from patients who reside in   **`r tools::toTitleCase(params$PatientAddressCounty)`** county or counties.

This is a summary table which includes sample metadata and mapping of the sequencing ID to the corresponding WDRS CASE_ID, if available
```{r overview_new_samples_table, echo = FALSE}

#Select columns based on epiprogram
if (params$epiprogram == "HAI") {
  summary_table_display<-summary_table_display %>% 
    select("WA ID",
    "ALT ID",
    "WDRS ID",
    "Collection Date",
    "Patient County",
    "Submitter County",
    "Submitter Facility",
    "Origin Facility")
} else {
   summary_table_display<-summary_table_display %>% 
    select("WA ID",
    "ALT ID",
    "WDRS ID",
    "Collection Date",
    "Patient County",
    "Submitter County",
    "Submitter Facility")
}
  
#Table attributes
kable(summary_table_display) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left",
    font_size = 12,
    html_font = "Arial"
  ) %>%
  row_spec(0, bold = TRUE)

```

The new sequencing results from patients who reside in   **`r tools::toTitleCase(params$PatientAddressCounty)`** county or counties were analyzed using **`r tools::toTitleCase(params$phylogeny)`** and classified by BigBacter as follows:
```{r new_isolates_table, echo = FALSE}
# Load info from new sequences
outputs_script_dir <- "outputs_scripts"

newseq_file <- list.files(outputs_script_dir, pattern = "newseq.RData$", full.names = TRUE)
loaded_newseq <- load(newseq_file[1])
summary_new_seq <- bind_rows(get(loaded_newseq[1]))


# Load partition_info
partition_file <- list.files(outputs_script_dir, pattern = "partition_info.RData$", full.names = TRUE)
loaded_partition <- load(partition_file[1])
partition_info <- get(loaded_partition[1])

# Filter and rename partition columns based on phylogeny parameter
if (params$phylogeny == "Snippy") {
  partition_info_filtered <- partition_info %>%
    select(ID_clean, PARTITION_Snippy, PARTITION_Snippy_Counts) %>%
    rename(
      PARTITION = PARTITION_Snippy,
      'PARTITION (n=)' = PARTITION_Snippy_Counts
    )
} else if (params$phylogeny == "Gubbins") {
  partition_info_filtered <- partition_info %>%
    select(ID_clean, PARTITION_Gubbins, PARTITION_Gubbins_Counts) %>%
    rename(
      PARTITION = PARTITION_Gubbins,
      'PARTITION (n=)' = PARTITION_Gubbins_Counts
    )
} else if (params$phylogeny == "all") {
  # Keep both Snippy and Gubbins columns
  partition_info_filtered <- partition_info %>%
    select(ID_clean, PARTITION_Snippy, PARTITION_Snippy_Counts,
           PARTITION_Gubbins, PARTITION_Gubbins_Counts) %>%
    rename(
      'PARTITION Snippy' = PARTITION_Snippy,
      'PARTITION Gubbins'= PARTITION_Gubbins,
      'PARTITION Snippy (n=)' = PARTITION_Snippy_Counts,
      'PARTITION Gubbins (n=)'= PARTITION_Gubbins_Counts
    )
} else {
  stop("Invalid phylogeny parameter. Choose 'Snippy', 'Gubbins', or 'all'.")
}

# Join partition info to summary_new_seq
summary_new_seq_partition <- summary_new_seq %>%
  left_join(partition_info_filtered, by = c("ID" = "ID_clean")) %>%
   rename(
     'GENOMIC CLUSTER' = GENOMIC_CLUSTER,
      'CLUSTER (n=)' = ISO_IN_CLUSTER)

#Using global helper function filter based on params PatientAddressCounty 
summary_new_seq <- summary_new_seq_partition %>%
  add_and_filter_county() %>%
  select(-any_of("PatientAddressCounty"))
  

# Table rendering
kable(summary_new_seq) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left",
    font_size = 12,
    html_font = "Arial"
  ) %>%
  row_spec(0, bold = TRUE)

```
*Sequences that resulted in new genetic clusters are excluded from tree partitioning.*

Of these, the following isolates resulted in new genetic clusters:

```{r new_clusters_table, echo = FALSE}
#Identify samples that resulted in new clusters
new_clust<-newseq %>% 
  filter(ISO_IN_CLUSTER==1)%>%
   rename(
      'CLUSTER (n=)' = ISO_IN_CLUSTER)

#Using global helper function filter based on params PatientAddressCounty 
new_clust <- new_clust %>%
  add_and_filter_county()


#Table attributes
if (nrow(new_clust) > 0) {
  kable(new_clust) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left",
    font_size = 12,
    html_font = "Arial"
  ) %>%
  row_spec(0, bold = TRUE)
} else {
  message("There were no new clusters created in this run")
}
```

## Failed Isolates

The following isolates failed quality control.

```{r failed_isolates_table, echo = FALSE}
outputs_script_dir <- "outputs_scripts"

failed_file <- list.files(outputs_script_dir, pattern = "failed.RData$", full.names = TRUE)
loaded_failed <- load(failed_file[1])
failed_df <- bind_rows(get(loaded_failed[1]))

#Using global helper function filter based on params PatientAddressCounty 
failed_df <- failed_df %>%
  add_and_filter_county()

if (nrow(failed_df) > 0) {
  kable(failed_df) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width = FALSE,
      position = "left",
      font_size = 12,
      html_font = "Arial"
    ) %>%
    row_spec(0, bold = TRUE)
} else {
  message("There weren't any isolates that failed in this run")
}

```

## Recombination

Bacterial recombination is the process where bacteria exchange genetic material with each other which leads to the gain of new DNA sequences into their genomes.  It is important to be aware of recombination when conducting genomic analyses because recombination events can be confused with mutations events which can impact metrics used to characterize relationships between sequences, such as calculating single nucleotide polymorphisms (SNP) distances. The bioinformatics pipelines developed at WA PHL use Gubbins, a method to detect and control for recombination. If recombination is detected the sites where recombination is present are masked in the SNPs distance calculations and in the phylogenetic trees. 

We evaluate recombination in multiple ways. First the number of sites where recombination was detected is divided by the total length of the core genome. If recombination is more than 5% in a genomic cluster the Gubbins outputs are used. If recombination is more than 1% but less than 5%, then the Snippy and Gubbins outputs are reviewed jointly to see if they yield different interpretations. If the interpretations differ, then most likely we will use the Gubbins for the genomic interpretations.

```{r recombination, echo = FALSE}
outputs_script_dir <- "outputs_scripts"

recomb_file <- list.files(outputs_script_dir, pattern = "recomb_summary.RData$", full.names = TRUE)

loaded_summary_recomb <- load(recomb_file[1])

summary_recomb <- bind_rows(get(loaded_summary_recomb[1]))


if (nrow(summary_recomb) > 0) {
  kable(summary_recomb) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width = FALSE,
      position = "left",
      font_size = 12,
      html_font = "Arial"
    ) %>%
    row_spec(0, bold = TRUE)
} else {
  message("Recombination wasn't calculated")
}

```

*Sequences that resulted in new genetic clusters are excluded from this calculation.*

## SNP Min and Max Distances

The minimum and maximum SNP distances calculated using **`r tools::toTitleCase(params$phylogeny)`** are summarized below.

```{r snp_min_max_table, echo = FALSE}
outputs_script_dir <- "outputs_scripts"

snpminmax_file <- list.files(outputs_script_dir, pattern = "summary_snp_minmax.RData$", full.names = TRUE)


loaded_snp_minmax <- load(snpminmax_file[1])

summary_snp <- bind_rows(get(loaded_snp_minmax[1]))

if (nrow(summary_snp) == 0) {

  cat("SNP distances were not calculated because this genomic cluster contains only a single sequence")

} else {

# Filter rows based on selected phylogeny (all, Snippy, or Gubbins)
if (params$phylogeny %in% c("Snippy", "Gubbins")) {
  summary_snp <- summary_snp %>%
    filter(grepl(params$phylogeny, Source, ignore.case = TRUE))
} else if (params$phylogeny == "all") {
  # Keep all rows
  summary_snp <- summary_snp
} else {
  stop("Invalid phylogeny parameter. Choose 'Snippy', 'Gubbins', or 'all'.")
}


if (nrow(summary_snp) > 0) {
  kable(summary_snp) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width = FALSE,
      position = "left",
      font_size = 12,
      html_font = "Arial"
    ) %>%
    row_spec(0, bold = TRUE)
} else {
  message("SNP distances were not calculated because this genomic cluster contains only a single sequence")
}
  
}

```
*Sequences that resulted in new genetic clusters are excluded from this calculation.*


## Genomic Linkages

Based on the SNP distances calculated using **`r tools::toTitleCase(params$phylogeny)`**, the following **very strong (0–5 SNPs)**, **strong (6–10 SNPs)**, and **intermediate (11–50 SNPs)** genomic linkages were identified between the new isolate(s) and other sequences within the corresponding genomic clusters.

```{r genomic_linkages_tables, echo=FALSE, results='asis'}

outputs_script_dir <- "outputs_scripts"


snplinks_file <- list.files(outputs_script_dir, pattern = "summary_snp_linkages.RData$", full.names = TRUE)
loaded_snp_links <- load(snplinks_file[1])
summary_snp_links <- bind_rows(get(loaded_snp_links[1]))

if (nrow(summary_snp_links) == 0) {
  cat("SNP distances weren't calculated\n")
} else {

  summary_snp_links <- summary_snp_links %>% filter(STATUS == "NEW")
  
  if (nrow(summary_snp_links) == 0) {

    cat("Genomic linkages were not detected among the new sequence/s and the other isolates in that genomic cluster\n")
  } else {
    
   # Filter by phylogeny parameter
    if (params$phylogeny %in% c("Snippy", "Gubbins")) {
      summary_snp_links <- summary_snp_links %>%
        filter(grepl(params$phylogeny, Source, ignore.case = TRUE))
    } else if (params$phylogeny == "all") {
      # keep all rows
      summary_snp_links <- summary_snp_links
    } else {
      stop("Invalid phylogeny parameter. Choose 'Snippy', 'Gubbins', or 'all'.")
    }
    
    if (nrow(summary_snp_links) == 0) {
      cat(paste0("No genomic linkages found for phylogeny: ", tools::toTitleCase(params$phylogeny), "\n"))
    } else {
      
    unique_sources <- unique(summary_snp_links$Source)

    for (source in unique_sources) {
      filtered_links <- summary_snp_links %>%
        filter(Source == source)

      if (nrow(filtered_links) > 0) {
        filtered_links <- filtered_links %>%
          select(-Source, -STATUS, -ID2, -dist) %>%
          rename(
            ID = ID1,
            `VeryStrongGenLinks (0-5 SNP)` = VeryStrongGenLinkage,
            `StrongGenLinks (6-10 SNP)` = StrongGenLinkage,
            `InterGenLinks (11-50 SNP)` = InterGenLinkage
          ) %>% 
         group_by(ID) %>%
         summarize(
          `VeryStrongGenLinks (0-5 SNP)` = str_remove(str_replace_all(paste(na.omit(unique(`VeryStrongGenLinks (0-5 SNP)`)), collapse = ", "), "\\s*,\\s*,", ", "), "^,\\s*"),
          `StrongGenLinks (6-10 SNP)` = str_remove(str_replace_all(paste(na.omit(unique(`StrongGenLinks (6-10 SNP)`)), collapse = ", "), "\\s*,\\s*,", ", "), "^,\\s*"),
`InterGenLinks (11-50 SNP)` = str_remove(str_replace_all(paste(na.omit(unique(`InterGenLinks (11-50 SNP)`)), collapse = ", "), "\\s*,\\s*,", ", "), "^,\\s*")
    ) %>%
      ungroup()
        
        cat(paste0("#### Genomic linkages for ", source, "\n\n"))

        cat(kable(filtered_links) %>%
          kable_styling(
            bootstrap_options = c("striped", "hover", "condensed", "responsive"),
            full_width = FALSE,
            position = "left",
            font_size = 12,
            html_font = "Arial"
          ) %>%
          row_spec(0, bold = TRUE) %>%
          as.character(), "\n")
        }
      }
    }
  }
}
```


## Metadata

This is an overview of the metadata pertaining to each of the genomic clusters that contain new isolates. The counties are extracted from patient county address.

```{r metadata_table, echo = FALSE}

results_dir <- "metadata_summ"

rdata_files <- list.files(results_dir, pattern = "\\.RData$", full.names = TRUE)


rdata_files <- rdata_files[!grepl("_sameDOB\\.RData$", rdata_files)]

#Extract taxa from the filename
extract_taxa <- function(filename) {
  parts <- unlist(strsplit(filename, "_"))  # Split by underscore
  return(paste(parts[1], parts[2], sep = "_"))  # Combine first two parts
}

#Empty list to store all data frames
all_data <- list()

#Load and collect each data frame
for (file in rdata_files) {
  load(file)
  file_name <- file_path_sans_ext(basename(file))
  taxa_name <- extract_taxa(file_name)
  
  #Add a new column to the data frame with the file name
  combined_df <- combined_df %>%
    mutate(Taxa_GenomicCluster = file_name)%>%
    filter(taxa_name == params$taxa)%>%
    select(Taxa_GenomicCluster, everything())
  
  #Store the modified data frame in the list
  all_data[[file_name]] <- combined_df
}

#Bind all the data frames into one combined data frame
final_metadata <- bind_rows(all_data) %>% 
  select(-"All_Facilities",
         -"New_Facilities",
         -"New_Counties",
         -"All_IDs",
         -"New_IDs")

#Table attributes
kable(final_metadata) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left",
    font_size = 10,
    html_font = "Arial"
  ) %>%
  row_spec(0, bold = TRUE)

```

### Isolates with the same DOB

These isolates share the same date of birth, which may indicate they originate from the same case. Verify against epi data to confirm whether the isolates listed are truly from the same case and consider this when interpreting the genomic data.

```{r samedob_table, echo = FALSE}

results_dir <- "metadata_summ"

rdata_files <- list.files(
  results_dir,
  pattern = "_sameDOB\\.RData$",
  full.names = TRUE
)

extract_taxa <- function(filename) {
  parts <- unlist(strsplit(filename, "_"))
  paste(parts[1], parts[2], sep = "_")
}

all_data <- list()

for (file in rdata_files) {
  load(file) 
  
  file_name <- file_path_sans_ext(basename(file))
  taxa_name <- extract_taxa(file_name)
  
  if (taxa_name == params$taxa) {
    
    temp_df <- same_dob_df %>%
      mutate(Taxa_GenomicCluster = file_name) %>%
      select(Taxa_GenomicCluster, everything())
    
    temp_df <- temp_df %>%
      add_and_filter_county()
    
    all_data[[file_name]] <- temp_df
  }
}

final_metadata <- bind_rows(all_data) %>%
  mutate(Taxa_GenomicCluster = gsub("_sameDOB$", "", Taxa_GenomicCluster)) %>%
  arrange(Taxa_GenomicCluster, PatientBirthDate, SpecimenDateCollected)

#Table attributes
if (nrow(final_metadata) > 0) {kable(final_metadata) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left",
    font_size = 10,
    html_font = "Arial"
  ) %>%
  row_spec(0, bold = TRUE)
} else {
  message("All isolates have unique dates of birth")
}
```


```{r draft_email, include=FALSE}

# ============================================================
# Draft email
# ============================================================

# Basic information
sequence_count <- nrow(summary_table_filtered)
pathogen_name <- params$taxa
phylogeny_name <- tools::toTitleCase(params$phylogeny)

# Start email
email_lines <- c(
  "Hello,",
  "",
  paste0(
    "There are ", sequence_count,
    " new sequence results for ", pathogen_name, "."
  ),
  "",
  paste0(
    "The ", phylogeny_name,
    " outputs were used for the following interpretations:"
  ),
  ""
)

# ------------------------------------------------------------
# Genomic cluster interpretations
# ------------------------------------------------------------

# Get unique genomic clusters represented among the new sequences
genomic_clusters <- unique(summary_new_seq$`GENOMIC CLUSTER`)
genomic_clusters <- genomic_clusters[!is.na(genomic_clusters)]

for (cluster in genomic_clusters) {

  # New sequences in this genomic cluster
  cluster_new <- summary_new_seq %>%
    filter(`GENOMIC CLUSTER` == cluster)

  # Number of isolates in genomic cluster
  cluster_size <- cluster_new$`CLUSTER (n=)`[1]

  # ----------------------------------------------------------
  # Genomic cluster heading
  # ----------------------------------------------------------

  email_lines <- c(
    email_lines,
    paste0(
      "Genomic Cluster ", cluster,
      " (n=", cluster_size, ")"
    ),
    ""
  )

  # ----------------------------------------------------------
  # Check whether this is a newly formed genomic cluster
  # ----------------------------------------------------------

  new_cluster <- new_clust %>%
    filter(GENOMIC_CLUSTER == cluster)

  if (nrow(new_cluster) > 0) {

    # This is a newly formed genomic cluster.
    # List each new sequence in the cluster.
    new_ids <- cluster_new$ID

    for (new_id in new_ids) {

      email_lines <- c(
        email_lines,
        paste0(
          "The new sequence ", new_id,
          " is in a newly formed genomic cluster; there are no genomic linkages."
        )
      )
    }

    email_lines <- c(email_lines, "")

    next
  }

  # ----------------------------------------------------------
  # Get unique partitions within this genomic cluster
  # ----------------------------------------------------------

  partitions <- unique(cluster_new$PARTITION)
  partitions <- partitions[!is.na(partitions)]

  for (partition in partitions) {

    # New sequences in this partition
    partition_new <- cluster_new %>%
      filter(PARTITION == partition)

    # Number of isolates in partition
    partition_size <- partition_new$`PARTITION (n=)`[1]

    # IDs in this partition
    new_ids <- partition_new$ID

    # --------------------------------------------------------
    # Partition heading
    # --------------------------------------------------------

    email_lines <- c(
      email_lines,
      paste0(
        "Partition ", partition,
        " (n=", partition_size, ")"
      ),
      ""
    )

    # --------------------------------------------------------
    # Get genomic linkages for new sequences in this partition
    # --------------------------------------------------------

    cluster_links <- summary_snp_links %>%
      filter(
        STATUS == "NEW",
        ID1 %in% new_ids | ID2 %in% new_ids
      )

    # --------------------------------------------------------
    # No Very Strong, Strong, or Intermediate linkages
    # --------------------------------------------------------

    if (nrow(cluster_links) == 0) {

      for (new_id in new_ids) {

        email_lines <- c(
          email_lines,
          paste0(
            "The new sequence ", new_id,
            " has no Very Strong, Strong, or Intermediate genomic linkages."
          )
        )
      }

      email_lines <- c(email_lines, "")

      next
    }

    # --------------------------------------------------------
    # Identify linkage category for each row
    # --------------------------------------------------------

    cluster_links <- cluster_links %>%
      mutate(
        new_id = if_else(ID1 %in% new_ids, ID1, ID2),
        linked_id = if_else(ID1 %in% new_ids, ID2, ID1),

        linkage_description = case_when(

          !is.na(VeryStrongGenLinkage) &
            trimws(VeryStrongGenLinkage) != "" ~
            "very close genomic linkage",

          !is.na(StrongGenLinkage) &
            trimws(StrongGenLinkage) != "" ~
            "close genomic linkage",

          !is.na(InterGenLinkage) &
            trimws(InterGenLinkage) != "" ~
            "intermediate genomic linkage",

          TRUE ~ NA_character_
        )
      )

    # --------------------------------------------------------
    # Report Very Strong and Strong linkages individually
    # --------------------------------------------------------

    individual_links <- cluster_links %>%
      filter(
        linkage_description %in% c(
          "very close genomic linkage",
          "close genomic linkage"
        )
      )

    if (nrow(individual_links) > 0) {

      for (i in seq_len(nrow(individual_links))) {

        link <- individual_links[i, ]

        # Get details for the new sequence
        new_sequence_details <- summary_table_display %>%
          filter(`WA ID` == link$new_id)

        # Select facility based on epidemiologic program
        if (params$epiprogram == "HAI") {

          facility <- new_sequence_details$`Origin Facility`
          facility_text <- paste0("at ", facility)

        } else {

          facility <- new_sequence_details$`Submitter Facility`
          facility_text <- paste0("submitted by ", facility)

        }

        email_lines <- c(
          email_lines,
          paste0(
            "The new sequence ", link$new_id,
            " collected on ", new_sequence_details$`Collection Date`,
            " from a case who resides in ",
            new_sequence_details$`Patient County`,
            " and was ", facility_text,
            " has a ", link$linkage_description,
            " (", link$dist, " SNPs) to ",
            link$linked_id, "."
          )
        )
      }
    }

    # --------------------------------------------------------
    # Count Intermediate linkages
    # --------------------------------------------------------

    intermediate_links <- cluster_links %>%
      filter(
        linkage_description == "intermediate genomic linkage"
      )

    if (nrow(intermediate_links) > 0) {

      # Count intermediate linkages by new sequence
      intermediate_counts <- intermediate_links %>%
        count(new_id, name = "n_intermediate")

      for (i in seq_len(nrow(intermediate_counts))) {

        new_id <- intermediate_counts$new_id[i]

        # Get details for the new sequence
        new_sequence_details <- summary_table_display %>%
          filter(`WA ID` == new_id)

        # Select facility based on epidemiologic program
        if (params$epiprogram == "HAI") {

          facility <- new_sequence_details$`Origin Facility`
          facility_text <- paste0("at ", facility)

        } else {

          facility <- new_sequence_details$`Submitter Facility`
          facility_text <- paste0("submitted by ", facility)

        }

        email_lines <- c(
          email_lines,
          paste0(
            "The new sequence ", new_id,
            " collected on ", new_sequence_details$`Collection Date`,
            " from a case who resides in ",
            new_sequence_details$`Patient County`,
            " and was ", facility_text,
            " has ",
            intermediate_counts$n_intermediate[i],
            " intermediate genomic ",
            ifelse(
              intermediate_counts$n_intermediate[i] == 1,
              "linkage",
              "linkages"
            ),
            ". See the attachment for more details."
          )
        )
      }
    }

    # Blank line between partitions
    email_lines <- c(email_lines, "")
  }
}

# ------------------------------------------------------------
# Write email to text file
# ------------------------------------------------------------

writeLines(
  email_lines,
  file.path(outputs_script_dir, "email_draft.txt")
)

```


## Resources
The code to generate this report is available here: \
<https://github.com/NW-PaGe/BacterialGenomicsSummaryOutput>\

The following bioinformatics methods were used by WA PHL to generate some of the data summarized in this report:\
BigBacter bioinformatics pipeline <https://github.com/doh-jdj0303/bigbacter-nf>\
Snippy <https://github.com/tseemann/snippy>\
Gubbins <https://github.com/nickjcroucher/gubbins>
